Reproducibility of SNV-calling in multiple sequencing runs from single tumors

### 2015-11-12 PAPER/paper.tex
edited page 2

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. To identify somatic mutations in cancer genomics data, researchers typically independently sequence the blood and tumor of a single patient and align the blood and tumor sequences to each other to identify somatic variants that are specific to the tumor DNA, any of which could be cancer-causing or -sustaining. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and variant-calling pipelines are not error-free. For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls. One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}. Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.

>>>>

Glioblastoma multiforme (GBM) is the most common and deadly primary brain tumor with a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. Recently, research to improve treatments for patients with this devastating disease has focused on the idea of precision medicine, that is using large-scale genomics data to discover how the disease arises and progresses, and how to stop it. The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}, including over 500 GBMs. While these and other similar data \citep{Parsons}, researchers have discovered genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.  Despite widespread use of the data and adoption of the methods, however, efforts to benchmark the data--to assess the quality and repeatability of these and subsequent analyses--have lagged or been non-existent. Here, our goal is to begin to develop robust methods for benchmarking next-generation cancer sequencing data and analysis.

It is well known that sequencing and variant-calling pipelines are not error-free. For example, different pipelines for calling single-nucleotide variants (SNVs) can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNV-calls. One way to address this question would be to analyze technical replicates of sequencing data \citep{replicates}. Here, we addressed one aspect of this question, by asking whether the same SNV-calling pipeline will return comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNV-calling increases or decreases the degree of similarity among replicates.
